A flexible integer linear programming formulation for scheduling clinician on-call service in hospitals

### 2019-08-30 paper/paper.intro.tex
Work on fixing discussion
- Remove/simplify discussion of NP-hardness
- Add more detail to impact of formulation on clinicians, hospital dept. w/citation
- Clean up ILP abbreviation throughout paper

Many of the approaches to nurse scheduling were designed to satisfy the requirements of a specific hospital department which causes a large number of variables and constraints to be incorporated into the problem formulation. While these department-specific approaches allow end-users to find precise schedules that satisfy the needs of the department and the preferences of the nurses and clinicians in that department, they are difficult to readily adapt to other departments in the same hospital or other hospitals. % explain why hard to adapt?... what makes their generalizablility/adaptability limited?
Moreover, the large number of variables and constraints also leads to computational complexity issues \cite{goos_complexity_1996}, especially when using exact methods for finding the solution. In this paper, we tackle a version of the nurse scheduling problem arising from a case study of one clinical division, providing two different services simultaneously (general infectious disease (ID) consults; and HIV consults service) at St. Michael's Hospital in Toronto, Canada. Our goal is to (1) present a simple integer linear programming formulations for the scheduling problem as developed for the hospital, and describe the adaptability of the formulation to solving similar problems in other departments; (2) compare the performance of the ILP scheduler to the results of the manual approach; and (3) analyze the robustness of the algorithm in difficult instances of the problem. \\
% present a simple formulation for the problem developed and tested at the hospital after switching from a manual approach to scheduling; and (2) analyze the performance of integer linear programming in solving difficult instances of the problem and compare the results with those of the manual approach; and (3) describe the adaptability of the formulation as a basic framework for solving similar problems in other departments. \\ %make #3 an objective

>>>>

Many of the approaches to nurse scheduling were designed to satisfy the requirements of a specific hospital department which causes a large number of variables and constraints to be incorporated into the problem formulation. While these department-specific approaches allow end-users to find precise schedules that satisfy the needs of the department and the preferences of the nurses and clinicians in that department, they are difficult to readily adapt to other departments in the same hospital or other hospitals. % explain why hard to adapt?... what makes their generalizablility/adaptability limited?
Moreover, the large number of variables and constraints also leads to computational complexity issues \cite{goos_complexity_1996}, especially when using exact methods for finding the solution. In this paper, we tackle a version of the nurse scheduling problem arising from a case study of one clinical division, providing two different services simultaneously (general infectious disease (ID) consults; and HIV consults service) at St. Michael's Hospital in Toronto, Canada. Our goal is to (1) present a simple integer linear programming (ILP) formulation for the scheduling problem as developed for the hospital, and describe the adaptability of the formulation to solving similar problems in other departments; (2) compare the performance of the ILP scheduler to the results of the manual approach; and (3) analyze the robustness of the algorithm in difficult instances of the problem. \\
% present a simple formulation for the problem developed and tested at the hospital after switching from a manual approach to scheduling; and (2) analyze the performance of integer linear programming in solving difficult instances of the problem and compare the results with those of the manual approach; and (3) describe the adaptability of the formulation as a basic framework for solving similar problems in other departments. \\ %make #3 an objective

### 2019-08-30 paper/paper.results.tex
Clean up abstract, intro, problem, methods, results
Make methods section more concise + add missing references
Add acks + funding sections
- Fix affiliation order, minor intro corrections
- TODO: add CPU specs to results section
Work on fixing discussion
- Remove/simplify discussion of NP-hardness
- Add more detail to impact of formulation on clinicians, hospital dept. w/citation
- Clean up ILP abbreviation throughout paper

\subsection{Software}
We developed a software package with a user interface that implements the above LP problem and allows configuration of clinicians at [ref \ref{???}], to be used by the ID division at St. Michael's Hospital. The software was used to generate the results in the following sections, using real data as well as simulated data as input.

\subsection{Infectious Diseases Division}  %main comment  = I think use past tense in writing
We used clinician time-off requests and minimum/maximum requirements from 2015-2018 as input data for the LP problem. Table \ref{tbl:2018-schedule-comparison} presents the optimal schedule generated using the software as well as the manually-created schedule for data from 2018. The schedule is color-coded to distinguish between the different clinicians assigned. \\

@@

First, we evaluated the software-generated schedule by comparing it with the manually-created schedule. Specifically, we examined the adherence of each schedule to the constraints presented in table \ref{tbl:constraint-summary}. As shown in table \ref{tbl:constraints-comparison}, the optimal schedule generated using the software satisfied all mandatory constraints. In contrast, the manually-created schedule was not able to satisfy all mandatory constraints. In particular, we see that the manual schedule assigned clinicians to multiple consecutive blocks in all four years. Moreover, the manual schedule did not have an equal distribution of weekends and holidays for all four years of data. Evaluating the objectives, we see that LP outperforms the manually created schedule in all four years, by accommodating almost all time-off requests and ensuring that weekends are always assigned close to blocks.

@@

\subsection{Simulations}
We then examined the properties of the LP problem using simulated data. Specifically, we examined the influence of expanding the following parameters on the runtime of the algorithm: number of clinicians; number of services offered in the division; amount of time-off requests per clinician throughout the year. We also examined the effect of a longer time-horizon on the runtime. \\

The effect of an increasing number of requests per clinician on the runtime of the LP solver is shown in figure \ref{fig:runtime-requests}. We executed the program on a department with 10 clinicians offering two services, similar to the department at St. Michael's Hospital. For simplicity, each clinician was configured with $\abs{\mathcal{BR}} = \{0, 5, 10, 15\}$ non-overlapping block requests. The increase in the number of requests did not affect the runtime of the LP solver, indicating that it can accommodate a lot of flexibility in clinician requests. Moreover, we see that all four runs were completed in under 1 second. \\

\begin{figure}[h]
	\centering
	\includegraphics[scale=.5]{fig/runtime_requests}
	\caption{Runtime of LP solver with an increasing number of requests per clinician}
	\label{fig:runtime-requests}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=.5]{fig/runtime_blocks}
	\caption{Runtime of LP solver with an increasing number of 2-week blocks}
	\label{fig:runtime-blocks}
\end{figure}

>>>>

\subsection{Software}
We developed a software package with a user interface that implements the above linear program and allows configuration of clinicians at [ref \ref{???}], to be used by the ID division at St. Michael's Hospital. The software was used to generate the results in the following sections, using real data as well as simulated data as input. All the following experiments were conducted on a [X-Core CPU] computer @ [Y GHz] and [Z GB] of RAM running 64bit Windows 10. The software was implemented using the COIN-OR Branch-and-Cut open source solver version 2.9.9 \cite{johnjforrest_coin-or/cbc:_2019}.

\subsection{Infectious Diseases Division}  %main comment  = I think use past tense in writing
We used clinician time-off requests and minimum/maximum requirements from 2015-2018 as input data for the ILP problem. Table \ref{tbl:2018-schedule-comparison} presents the optimal schedule generated using the software as well as the manually-created schedule for data from 2018. The schedule is color-coded to distinguish between the different clinicians assigned. \\

@@

First, we evaluated the software-generated schedule by comparing it with the manually-created schedule. Specifically, we examined the adherence of each schedule to the constraints presented in table \ref{tbl:constraint-summary}. As shown in table \ref{tbl:constraints-comparison}, the optimal schedule generated using the software satisfied all mandatory constraints. In contrast, the manually-created schedule was not able to satisfy all mandatory constraints. In particular, we see that the manual schedule assigned clinicians to multiple consecutive blocks in all four years. Moreover, the manual schedule did not have an equal distribution of weekends and holidays for all four years of data. Evaluating the objectives, we see that LP outperforms the manually created schedule in all four years, by accommodating almost all time-off requests and ensuring that weekends are always assigned close to blocks. \\

@@

\subsection{Simulations}
We then examined the properties of the ILP problem using simulated data. Specifically, we examined the influence of expanding the following parameters on the runtime of the algorithm: number of clinicians; number of services offered in the division; amount of time-off requests per clinician throughout the year. We also examined the effect of a longer time-horizon on the runtime. \\

The effect of an increasing number of requests per clinician on the runtime of the ILP solver is shown in figure \ref{fig:runtime-requests}. We executed the program on a department with 10 clinicians offering two services, similar to the department at St. Michael's Hospital. For simplicity, each clinician was configured with $\abs{\mathcal{BR}} = \{0, 5, 10, 15\}$ non-overlapping block requests. The increase in the number of requests did not affect the runtime of the ILP solver, indicating that it can accommodate a lot of flexibility in clinician requests. Moreover, we see that all four runs were completed in under 1 second. \\

\begin{figure}[h]
	\centering
	\includegraphics[scale=.5]{fig/runtime_requests}
	\caption{Runtime of ILP solver with an increasing number of requests per clinician}
	\label{fig:runtime-requests}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=.5]{fig/runtime_blocks}
	\caption{Runtime of ILP solver with an increasing number of 2-week blocks}
	\label{fig:runtime-blocks}
\end{figure}
